# Exploratory Data Analysis

## Early questions

-   does age have any impact on how a squirrel acts (kuks, qaas, indifferent, etc.)?
-   how does location influence the sounds or behavior of a squirrel?
-   which factors (**location**, time of day, **primary fur color**) most impact a squirrel's interaction with humans?
-   make plots- scatterplots
-   primary fur color: 1 = Gray, 2 = Cinnamon, 3 = Black

Loading in new cleaned dataset

```{r}
# Install necessary packages if not already installed
if (!require("dplyr")) install.packages("dplyr")
if (!require("janitor")) install.packages("janitor")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("tidyr")) install.packages("tidyr")
if (!require("sf")) install.packages("sf")
if (!require("mapview")) install.packages("mapview")
if (!require("maps")) install.packages("maps")
if (!require("OpenStreetMap")) install.packages("OpenStreetMap")
if (!require("caret")) install.packages("caret")
if (!require("e1071")) install.packages("e1071")
if (!require("plotly")) install.packages("plotly")
if (!require("gridExtra")) install.packages("gridExtra")
if (!require("readr")) install.packages("readr")

# Load necessary libraries
library(dplyr)
library(janitor)
library(ggplot2)
library(tidyr)
library(sf)
library(mapview)
library(maps)
library(OpenStreetMap)
library(caret)  # For cross-validation
library(e1071)  # For logistic regression
library(plotly)  # For 3D plots
library(gridExtra)
library(readr)

getwd()

filename <- "RaimaSaha/squirrel_data_clean.csv"

clean_data <- read.csv(filename)  

head(clean_data)
clean_data
```

Looking at the structure of the data

```{r}
str(clean_data)
```

Looking at the summary of the data

```{r}
summary(clean_data)
```

Comparing number of adult squirrels to juvenile squirrels

```{r}
age_column <- clean_data |>
  select(age) |>
  drop_na() # dropping NA values in the data set

age_column |>
  count(age) |>
  ggplot(aes(x = age, y = n, fill = age)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Adults versus Juveniles",
  x = "Age",
  y = "Count",
  fill = "Age")
```

I don't know what the pink (?) category. Take a look into the distribution of the data.

```{r}
table(clean_data$age) # distribution of adult and juvenile
sum(is.na(clean_data$age)) # number of NA values
```

Creating plot to see trend between sighting above or below ground and date.

```{r}
location_date_column <- clean_data |>
  select(location, date) |>
  drop_na() # dropping NA values in the data set

ggplot(location_date_column, aes(x = location, fill = factor(date))) + geom_bar(position = "dodge") 
```

We can see that squirrels are more likely to be found on the ground plane than above ground on any given day.

Let's look at the location of a squirrel based on its fur color.

```{r}
location_fur_column <- clean_data |>
  select(location, primary_fur_color) |>
  drop_na() # dropping NA values in the data set

ggplot(location_fur_column, aes(x = location, fill = primary_fur_color)) + geom_bar(position = "dodge") 
```

Not much can be discerned by the graph other than the trend discovered above, that squirrels tend to spotted more on the ground plane than above ground.

Let's create a plot to see if a certain colored squirrel is to be spotted more on any given day.

```{r}
fur_date_column <- clean_data |>
  select(primary_fur_color, date) |>
  drop_na() # dropping NA values in the data set

ggplot(fur_date_column, aes(x = primary_fur_color, fill = factor(date))) + geom_bar(position = "dodge") 
```

From this plot, we can see that squirrels with gray as their primary fur color are spotted more than squirrels with any other fur color. Maybe gray squirrels are the most common?

I'm creating a plot to see how high each squirrel was spotted above ground based on primary fur color. Knowing that majority of the squirrels in the data set have gray fur color and they were spotted above ground the most, I wouldn't be surprised if the highest measurements belonged to them.

```{r}
fur_agm_column <- clean_data |>
  select(primary_fur_color, above_ground_sighter_measurement) #|>
  #drop_na() # dropping NA values in the data set

ggplot(fur_agm_column, aes(x = primary_fur_color, y = above_ground_sighter_measurement)) + geom_bar(stat = "identity") 
```

My prediction was right.

**IGNORE, COME BACK TO THIS:** Let's look at the relationship between primary fur colors and latitudinal and longitudinal locations.

```{r}
# lat_long <- clean_data |>
#   select(latitude, longitude, primary_fur_color)
# 
# ggplot(data = lat_long, aes(map_id = city)) + geom_map(aes(fill = primary_fur_color, map = maps_data("city"))) 

  

upperLeft = c(40.75, -73.50)

lowerRight = c(40.71, -73.43)

base_map  = openmap(upperLeft, lowerRight, type="osm")

plot(base_map)

points <- st_as_sf(clean_data, coords = c("latitude", "longitude"), crs = 4326)

plot(st_geometry(points), pch = 16, col = "navy", cex = 2, add = T)

```

Creating plot mapping out latitudes and longitudes in mapview. Shows distribution of squirrels without any context.

```{r}
mapview(clean_data, xcol = "longitude", ycol = "latitude", crs = 4326, grid = FALSE)


```

Created a pie chart of number of squirrels found running and not running

```{r}
pie(table(clean_data$running), main = "Running?",  labels = c("False", "True"), sep = "")

table(clean_data$running)
```

Time of day vs primary fur color

```{r}
ggplot(clean_data, aes(x = shift, fill = primary_fur_color)) + geom_bar(position = "dodge")
```

Location vs primary fur color

```{r}
spec_data <- clean_data |>
  select(primary_fur_color, location) |>
  drop_na()

ggplot(spec_data, aes(x = primary_fur_color, y = location, color = primary_fur_color)) + geom_count() + labs(title = "Location vs. Primary Fur Color", x = "Primary Fur Color", y = "Location", size = "Count", color = "Primary Fur Color")
```

Kuks vs primary fur color

```{r}
spec_data <- clean_data |>
  select(primary_fur_color, kuks) |>
  drop_na()

ggplot(spec_data, aes(x = primary_fur_color, y = kuks, color = primary_fur_color)) + geom_count() + labs(title = "Kuks vs. Primary Fur Color", x = "Primary Fur Color", y = "Kuks", size = "Count", color = "Primary Fur Color")
```

foraging vs primary fur color

```{r}
spec_data <- clean_data |>
  select(primary_fur_color, foraging) |>
  drop_na()

ggplot(spec_data, aes(x = primary_fur_color, y = foraging, color = primary_fur_color)) + geom_count() + labs(title = "Foraging vs. Primary Fur Color", x = "Primary Fur Color", y = "Foraging", size = "Count", color = "Primary Fur Color")
```

Looking at the distributions of adults and juveniles by latitudinal and longitudinal coordinates.

```{r}
my_sf <- clean_data |>
  st_as_sf(coords = c('longitude', 'latitude')) |>
  st_set_crs(4326)

ggplot() + geom_sf(data = my_sf, aes(color = age)) + theme(axis.text.x = element_text(angle = 90))
```

Looking at squirrel location in relation to hectare, and also to foraging

```{r}
my_sf <- clean_data |>
  st_as_sf(coords = c('longitude', 'latitude')) |>
  st_set_crs(4326)

ggplot() + geom_sf(data = my_sf, aes(color = hectare, shape = factor(foraging))) + theme(axis.text.x = element_text(angle = 90))
```

Looking at latitudinal/longitudinal coordinates in relation to primary fur color

```{r}
my_sf <- clean_data |>
  st_as_sf(coords = c('longitude', 'latitude')) |>
  st_set_crs(4326)

ggplot() + geom_sf(data = my_sf, aes(color = primary_fur_color)) + theme(axis.text.x = element_text(angle = 90))  
```

Looking at relationship between geographical location, primary fur color, and age.

```{r}
my_sf <- clean_data |>
  filter(!is.na("longitude") & !is.na("latitude")) |>
  st_as_sf(coords = c('longitude', 'latitude')) |>
  st_set_crs(4326)

ggplot() + geom_sf(data = my_sf, aes(color = primary_fur_color, shape = age), size = 2) + theme(axis.text.x = element_text(angle = 90)) +  facet_wrap(~primary_fur_color) 

ggplot() + geom_sf(data = my_sf, aes(color = primary_fur_color, shape = age), size = 2) + theme(axis.text.x = element_text(angle = 90))

summary(clean_data)
```

Looking at the date of the observations made vs. location in the park, also looking at age.

```{r}
my_sf <- clean_data |>
  st_as_sf(coords = c('longitude', 'latitude')) |>
  st_set_crs(4326)

ggplot() + geom_sf(data = my_sf, aes(color = date, shape = age)) + theme(axis.text.x = element_text(angle = 90))
```
