# Principal Component Analysis

Question: What factors impact squirrel's behavior around humans (approaches, runs from, or indifferent)?

Loading in data

```{r}
# Install necessary packages if not already installed
if (!require("dplyr")) install.packages("dplyr")
if (!require("janitor")) install.packages("janitor")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("tidyr")) install.packages("tidyr")
if (!require("sf")) install.packages("sf") 
if (!require("mapview")) install.packages("mapview")
if (!require("maps")) install.packages("maps")
if (!require("OpenStreetMap")) install.packages("OpenStreetMap")
if (!require("caret")) install.packages("caret")
if (!require("e1071")) install.packages("e1071")
if (!require("plotly")) install.packages("plotly")
if (!require("gridExtra")) install.packages("gridExtra")
if (!require("readr")) install.packages("readr")

# Load necessary libraries
library(dplyr)
library(janitor)
library(ggplot2)
library(tidyr)
library(sf)
library(mapview)
library(maps)
library(OpenStreetMap)
library(caret)  # For cross-validation
library(e1071)  # For logistic regression
library(plotly)  # For 3D plots
library(gridExtra)
library(readr)

getwd()

filename <- "RaimaSaha/squirrel_data_clean.csv"

clean_data <- read.csv(filename)  

head(clean_data)
clean_data
```

Conducting PCA

```{r}
# Standardize the data for running:foraging
temp_data <- clean_data |>
  select(latitude, longitude, primary_fur_color_num, shift_num, location_num) |>
  drop_na()
scaled_data <- scale(temp_data[, ])

# Perform PCA
pca <- prcomp(scaled_data, center = TRUE, scale. = TRUE)

# Get covariance matrix and eigenvalues
cov_matrix <- cov(scaled_data)
eigenvalues <- pca$sdev^2
explained_variance_ratio <- eigenvalues / sum(eigenvalues)

print("Covariance Matrix:")
print(cov_matrix)
print("\nEigenvalues:")
print(eigenvalues)

# corrMatrix = np.corrcoef(x,y) #input values for each of the variables
# # Plot the data:
# plt.imshow(corrMatrix) 
# plt.xlabel(‘x’)
# plt.ylabel(‘y’)
# plt.colorbar()
# plt.show()
```

Explained variance ratio

```{r}
# Explained Variance Ratio
total_explained_variances <- cumsum(explained_variance_ratio)

print("Explained Variance Ratios:")
print(explained_variance_ratio)

for (i in 1:length(explained_variance_ratio)) {
  cat('Total explained variance including', i, 'components:', total_explained_variances[i], '\n')
  cat('We', ifelse(total_explained_variances[i] < 0.9, 'should use more components', 'do not need to use more components.'), '\n')
}
```

Creating a scree plot

```{r}
scree_plot <- ggplot(data.frame(PC = 1:length(eigenvalues), Eigenvalue = eigenvalues), aes(x = PC, y = Eigenvalue)) +
  geom_line() +
  geom_point() +
  xlab('Principal Component') +
  ylab('Eigenvalue') +
  ggtitle('Scree Plot') +
  theme_minimal()

print(scree_plot)
```

```{r}
variance_df <- data.frame(PC = 1:length(explained_variance_ratio), Variance = explained_variance_ratio, Cumulative = total_explained_variances)
variance_plot <- ggplot(variance_df, aes(x = PC)) +
  geom_bar(aes(y = Variance), stat = 'identity', alpha = 0.5, fill = 'blue') +
  geom_step(aes(y = Cumulative), color = 'red') +
  xlab('Principal Components') +
  ylab('Variance Explained') +
  ggtitle('Explained Variance Ratio and Cumulative Explained Variance') +
  theme_minimal()

print(variance_plot)
```

```{r}
print("Explained Variances:")
print(eigenvalues)

for (i in 1:length(eigenvalues)) {
  ev <- eigenvalues[i]
  cat('Component', i, ' has explained variance of:', ev, '\n')
  cat('We will', ifelse(ev <= 1, 'not', ''), 'use this component.\n')
}
```

```{r}
# Visualize the top 2 components of the data
# pc_1 <- pca$x[, 1]
# pc_2 <- pca$x[, 2]
# 
# qplot(pc_1, pc_2, color = as.factor(data$Class)) +
#   xlab('Principal Component 1') +
#   ylab('Principal Component 2') +
#   ggtitle('PCA of Artificial Data') +
#   theme_minimal()

autoplot(pca, scaled_data, color = "primary_fur_color_num") +
  xlab('Principal Component 1') +
  ylab('Principal Component 2') +
  ggtitle('PCA of Data') +
  theme_minimal()
```

Better understanding of the linear transformation of the features

```{r}
biplot(pca)
```

Plotting the PCA

```{r}
plot(pca, type = "bar")
plot(pca, type = "lines")
```

3D PCA plot

```{r}
pca
components <- pca[["x"]]
components <- data.frame(components)
components$PC3 <- -components$PC3
components$PC2 <- -components$PC2
components = cbind(components, temp_data$primary_fur_color)

total_explained_variance_ratio <- summary(pca)[["importance"]]["Proportion of Variance",]
total_explained_variance_ratio <- 100 * sum(total_explained_variance_ratio)
total_explained_variance_ratio

fig <- plot_ly(components, x = ~PC1, y = ~PC2, z = ~PC3, color = ~temp_data$primary_fur_color, colors = c('black','brown','gray') ) |>
  add_markers(size = 12)
  
fig <- fig |>
  layout(
    title = cat("Total Explained Variance = ", total_explained_variance_ratio),
    scene = list(bgcolor = "#e5ecf6")
)

fig
```

Comparing each PC t each other

```{r}
components <- pca[["x"]]
components <- data.frame(components)
components <- cbind(components, temp_data$primary_fur_color)
# components$PC3 <- -components$PC3
components$PC2 <- -components$PC2

axis = list(showline=FALSE,
            zeroline=FALSE,
            gridcolor='#ffff',
            ticklen=4,
            titlefont=list(size=13))

fig <- components |>
  plot_ly()  |>
  add_trace(
    type = 'splom',
    dimensions = list(
      list(label=paste('PC 1 (',toString(round(explained_variance_ratio[1],1)),'%)',sep = ''), values=~PC1),
      list(label=paste('PC 2 (',toString(round(explained_variance_ratio[2],1)),'%)',sep = ''), values=~PC2),
      list(label=paste('PC 3 (',toString(round(explained_variance_ratio[3],1)),'%)',sep = ''), values=~PC3),
      list(label=paste('PC 4 (',toString(round(explained_variance_ratio[4],1)),'%)',sep = ''), values=~PC4),
      list(label=paste('PC 5 (',toString(round(explained_variance_ratio[5],1)),'%)',sep = ''), values=~PC5)#,
      # list(label=paste('PC 6 (',toString(round(explained_variance_ratio[6],1)),'%)',sep = ''), values=~PC6),
      # list(label=paste('PC 7 (',toString(round(explained_variance_ratio[7],1)),'%)',sep = ''), values=~PC7),
      # list(label=paste('PC 8 (',toString(round(explained_variance_ratio[8],1)),'%)',sep = ''), values=~PC8)
    ),
    color = ~temp_data$primary_fur_color, colors = c('black','brown','gray')
  ) %>%
  style(diagonal = list(visible = FALSE)) %>%
  layout(
    legend=list(title=list(text='color')),
    hovermode='closest',
    dragmode= 'select',
    plot_bgcolor='rgba(240,240,240, 0.95)',
    xaxis=list(domain=NULL, showline=F, zeroline=F, gridcolor='#ffff', ticklen=4),
    yaxis=list(domain=NULL, showline=F, zeroline=F, gridcolor='#ffff', ticklen=4),
    xaxis2=axis,
    xaxis3=axis,
    xaxis4=axis,
    yaxis2=axis,
    yaxis3=axis,
    yaxis4=axis
  )

fig
```

Visualizing top tow pricipal compnents in 2D

```{r}
fig <- plot_ly(components, x = ~PC1, y = ~PC2, color = ~temp_data$primary_fur_color, colors = c("black", "brown", "gray"), type = "scatter", mode = "markers") |>
  layout(
    legend = list(title = list(text='primary fur color')),
    plot_bgcolor = "#e5ecf6",
    xaxis = list(
      title = "PC1",
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor = "#ffff"),
    yaxis = list(
      title = "PC2",
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor = "#ffff"))
  fig
```
