# Chi-Squared Testing

Loading in data and packages

```{r}
# Install necessary packages if not already installed
if (!require("dplyr")) install.packages("dplyr") 
if (!require("janitor")) install.packages("janitor")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("tidyr")) install.packages("tidyr") 
if (!require("stringr")) install.packages("stringr")
if (!require("corrplot")) install.packages("corrplot")

# Load necessary libraries
library(tidyr) 
library(tidyverse) 
library(dplyr) 
library(ggplot2) 
library(stringr)
library(corrplot)

# Loading in data
getwd()

filename <- "RaimaSaha/squirrel_data_clean.csv"

clean_data <- read.csv(filename)  

head(clean_data)
clean_data
```

Null hypothesis: There is no significant relationship or correlation between variable \_\_\_\_\_ and human_interaction.

Alternative hypothesis: There is a significant relationship or correlation between variable \_\_\_\_\_\_ and human_interaction.

Setting significance level at 0.05 – if p-value is less than 0.05, then there is a correlation between \_\_\_\_\_\_\_\_ and human_interaction; otherwise, there isn't.

# Hectare and Human Interaction

Conducting chi-squared analysis between hectare and human interaction

```{r}
# creating a contingency table
contingency_table <- table(clean_data$hectare, clean_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)
```

Extracting results

```{r}
# extracting p-value
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals
residuals
```

## Hectare_num

Too few data points per each value occurrence is causing chi-squared test to be faulty and not produce a p-value. Creating a new version of the hectare column where it will only grouped by numbers. For example, entries that correspond with 42A-42C will be grouped under 42.

```{r}
clean_data
hectare_group_data <- separate(clean_data, hectare, c("hectare_num", "hectare_letter"), sep = 2)
hectare_group_data
```

Conduction chi-squared analysis between hectare_num and human interaction

```{r}
# creating a contingency table
contingency_table <- table(hectare_group_data$hectare_num, hectare_group_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)
```

Extracting results

```{r}
# extracting p-value
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals
residuals
```

1.647426e-11 \< 0.05 ∴ there is a statistically significant association between hectare_num and human interaction.

Plotting correlogram

```{r}
corrplot(chisq_results$residuals, is.cor = FALSE, method = "color", title = "Correlation between Hectare and Human Interaction", mar=c(0,0,1,0))
```

## Hectare_tens

Separating hectare_num column into just tens place to see if chi-squared test can be more precise

```{r}
clean_data
hectare_group_data
hectare_tens_data <- separate(hectare_group_data, hectare_num, c("hectare_tens", "hectare_ones"), sep = 1)
hectare_tens_data
```

Conducting chi-squared analysis between hectare_tens and human interaction

```{r}
# creating a contingency table
contingency_table <- table(hectare_tens_data$hectare_tens, hectare_tens_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)
```

Extracting results

```{r}
# extracting p-value
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals
residuals
```

3.649006e-16 \< 0.05 ∴ there is a statistically significant association between hectare_tens and human interaction.

Plotting correlation plot

```{r}
corrplot(chisq_results$residuals, is.cor = FALSE, method = "color", title = "Correlation between Hectare Tens and Human Interaction", mar=c(0,0,1,0))
```

# Shift and Human interaction

Conducting chi-squared analysis between shift and human interaction

```{r}
# creating a contingency table
contingency_table <- table(clean_data$shift, clean_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)
```

Extracting results

```{r}
# extracting p-value
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals
residuals
```

0.005426958 \< 0.05 ∴ there is statistically significant association between shift and human interaction.

Correlogram

```{r}
corrplot(chisq_results$residuals, is.cor = FALSE, method = "color", title = "Correlation between Shift and Human Interaction", mar=c(0,0,1,0))
```

# Location and Human Interaction

Conducting chi-squared analysis between location and human interaction

```{r}
# creating a contingency table
contingency_table <- table(clean_data$location, clean_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)
```

Extracting results

```{r}
# extracting p-value
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals
residuals
```

0.008341746 \< 0.05 ∴ there is a statistically significant association between ground location and human interaction.

Correlogram

```{r}
corrplot(chisq_results$residuals, is.cor = FALSE, method = "color", title = "Correlation between Location and Human Interaction", mar=c(0,0,1,0))
```

# Primary fur color and Human interaction

Conducting chi-squared analysis between primary fur color and human interaction

```{r}
# creating a contingency table
contingency_table <- table(clean_data$primary_fur_color, clean_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)
```

Extracting results

```{r}
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals
residuals
```

6.31357e-06 \< 0.05 ∴ there is a statistically significant association between primary fur color and human interaction.

Correlogram

```{r}
corrplot(chisq_results$residuals, is.cor = FALSE, method = "color", title = "Correlation between Primary Fur Color and Human Interaction", mar=c(0,0,1,0))
```
