---
title: "Various factors found significant to predicting the Eastern gray squirrel in Central Park, NYC interaction with humans"
author: "Raima Saha"
format: html
editor: visual
---

## Abstract

# Background, Research Question, and Hypotheses

# Results

## Principle Component Analysis

```{r}
#! echo: false
#! message: false
# Install necessary packages if not already installed
if (!require("dplyr")) install.packages("dplyr")
if (!require("janitor")) install.packages("janitor")
# if (!require("ggplot2")) install.packages("ggplot2")
if (!require("tidyr")) install.packages("tidyr")
# if (!require("sf")) install.packages("sf") 
# if (!require("mapview")) install.packages("mapview")
# if (!require("maps")) install.packages("maps")
# if (!require("OpenStreetMap")) install.packages("OpenStreetMap")
if (!require("caret")) install.packages("caret")
if (!require("e1071")) install.packages("e1071")
if (!require("plotly")) install.packages("plotly")
if (!require("gridExtra")) install.packages("gridExtra")
if (!require("readr")) install.packages("readr")
if (!require("ggfortify")) install.packages("ggfortify")
if (!require("corrplot")) install.packages("corrplot")
if (!require("ggbiplot")) install.packages("ggbiplot")
if (!require("ggrepel")) install.packages("ggrepel")
# if (!require("stats")) install.packages("stats")

# Load necessary libraries
library(dplyr)
 library(janitor)
# library(ggplot2)
library(tidyr)
# library(sf)
# library(mapview)
# library(maps)
# library(OpenStreetMap)
library(caret)  # For cross-validation
library(e1071)  # For logistic regression
library(plotly)  # For 3D plots
library(gridExtra)
library(readr)
library(ggfortify)
library(corrplot)
library(ggbiplot)
library(ggrepel)
# library(stats)

getwd()

filename <- "/Users/raimasaha/DSRP-2024-Devyani/RaimaSaha/squirrel_data_clean.csv"

clean_data <- read.csv(filename)  
```

Table 1 displays a covariance matrix yielded after conducting PCA on the variables latitude, longitude, primary_fur_color_num, shift_num, and location_num.

```{r}
# Standardize the data for running:foraging
temp_data <- clean_data |>
  select(latitude, longitude, primary_fur_color_num, shift_num, location_num, human_interaction_num) |>
  drop_na()
scaled_data <- scale(temp_data[, ])

# Perform PCA
pca <- prcomp(scaled_data, center = TRUE, scale. = TRUE)

# Get covariance matrix and eigenvalues
cov_matrix <- cov(scaled_data)
cov_matrix
eigenvalues <- pca$sdev^2
explained_variance_ratio <- eigenvalues / sum(eigenvalues)
```

Positive values indicate a positive relationship between certain principle components, while negative values indicate a negative relationship between certain principle components.

Figure 1 returns a correlogram between the principle components in the covariance matrix.

```{r}
corrplot(cov_matrix, method = "color")
```

Table 2 shows the total explained variance. It is said that

```{r}
# Explained Variance Ratio
total_explained_variances <- cumsum(explained_variance_ratio)

# print("Explained Variance Ratios:")
# print(explained_variance_ratio)

explained_variance_ratio_vector <- c()
components_vector <- c()

for (i in 1:length(explained_variance_ratio)) {
  # cat('Total explained variance including', i, 'components:', total_explained_variances[i], '\n')
  # cat('We', ifelse(total_explained_variances[i] < 0.9, 'should use more components', 'do not need to use more components.'), '\n')
  components_vector <- append(components_vector, i)
  explained_variance_ratio_vector <- append(explained_variance_ratio_vector, total_explained_variances[i])
}
df <- data.frame(number_of_components = components_vector, total_explained_variance = explained_variance_ratio_vector)
df
```

Figure 2 depicts the explained variance ratio and their corresponding values of how much variance they explain.

```{r}
variance_df <- data.frame(PC = 1:length(explained_variance_ratio), Variance = explained_variance_ratio, Cumulative = total_explained_variances)
variance_plot <- ggplot(variance_df, aes(x = PC)) +
  geom_bar(aes(y = Variance), stat = 'identity', alpha = 0.5, fill = 'blue') +
  geom_step(aes(y = Cumulative), color = 'red') +
  xlab('Principal Components') +
  ylab('Variance Explained') +
  ggtitle('Explained Variance Ratio and Cumulative Explained Variance') +
  theme_minimal()

print(variance_plot)
```

Figure 3 shows a Scree plot of all five principal components. It is consistent with the Explained Variance Ration in showing that 4 components is enough explain majority of the variance within the chosen variables.

```{r}
scree_plot <- ggplot(data.frame(PC = 1:length(eigenvalues), Eigenvalue = eigenvalues), aes(x = PC, y = Eigenvalue)) +
  geom_line() +
  geom_point() +
  xlab('Principal Component') +
  ylab('Eigenvalue') +
  ggtitle('Scree Plot') +
  theme_minimal()

print(scree_plot)
```

Figure 4 is a biplot of the PCA. Latitude and longitude appear to be closely correlated. The near 90 degree angle by shift, location, and latitude and longitude vectors show that these variables are not closely correlated.

```{r}
ggbiplot(pca, obs.scale = 1, var.scale = 1, ellipse = TRUE, circle = TRUE) + scale_x_discrete(labels=c("shift_num"="Shift", "primary_fur_color_num"="Primary Fur Color")) + ggtitle("PC1 vs PC2") + labs(color = "Shift") + theme_minimal()
```

Figure 5 shows a 2D representation of principal component 1 and principal component 2.

```{r}
components <- pca[["x"]]
components <- data.frame(components)
components <- cbind(components, temp_data$primary_fur_color)
# components$PC3 <- -components$PC3
components$PC2 <- -components$PC2

axis = list(showline=FALSE,
            zeroline=FALSE,
            gridcolor='#ffff',
            ticklen=4,
            titlefont=list(size=13))

plot_ly(components, x = ~PC1, y = ~PC2, color = ~temp_data$primary_fur_color_num, colors = c("black", "brown", "gray"), type = "scatter", mode = "markers") |>
  layout(
    title = list(
      title = "PC 1 vs. PC 2"),
    plot_bgcolor = "#e5ecf6",
    xaxis = list(
      title = "PC1",
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor = "#ffff"),
    yaxis = list(
      title = "PC2",
      zerolinecolor = "#ffff",
      zerolinewidth = 2,
      gridcolor = "#ffff"),
    legend = list(
      title ="Primary Fur Color"))
```

## Chi-Squared Testing

After conducting a chi-squared test between hectare and human interaction, shift and human interaction, location and human interaction, and primary fur color and human interaction, it was found that all these variables had a p-value of less than 0.05, the set significance level, indicating that there is a significant correlation between of the two variable and to reject the null hypothesis.

### Hectare and Human Interaction

```{r}
# Separating Hectare information into tens, ones, and letter
hectare_group_data <- separate(clean_data, hectare, c("hectare_num", "hectare_letter"), sep = 2)
hectare_tens_data <- separate(hectare_group_data, hectare_num, c("hectare_tens", "hectare_ones"), sep = 1)

# creating a contingency table
contingency_table <- table(hectare_tens_data$hectare_tens, hectare_tens_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)

# extracting p-value
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals

```

3.649006e-16 is less than the set significance level of 0.05, meaning that there is a significant correlation between Hectare and Human Interaction.

Figure 6 below displays a correlogram showing that squirrels with a primary fur color of cinnamon are more likely to approach human, and that gray squirrels tend to have slight inclination to not approach humans.

```{r}
corrplot(chisq_results$residuals, is.cor = FALSE, method = "color", title = "Correlation between Hectare and Human Interaction", mar=c(0,0,1,0))
```

## Shift and Human Interaction

```{r}
# creating a contingency table
contingency_table <- table(clean_data$shift, clean_data$human_interaction)

# performing chi-squared test
chisq_results <- chisq.test(contingency_table)

# extracting p-value
p_val <- chisq_results$p.value
p_val

# extracting residuals
residuals <- chisq_results$residuals
```

0.005426958 is less than 0.05, thereby indicating a significant correlation between shift and human interaction.

Figure 7 below displays the correlation between shift and human interaction.

```{r}
corrplot(chisq_results$residuals, is.cor = FALSE, method = "color", title = "Correlation between Shift and Human Interaction", mar=c(0,0,1,0))
```

## Primary Fur Color and Human Interaction

# Discussion

Acknowledgements

Sources
